The five pillars of computational reproducibility

### 2023-05-08 docs/5pillars.Rmd
5pil update

Unreliable research is a waste of resources and time, and can mislead readers to pursue doomed
research directions.
Much has been written on how irreproducibility has reached crisis levels in psychology and life
sciences [@Baker2015-bs; @Begley2012-di].
Indeed in bioinformatics, the situation is grim with only a very fraction of studies
being able to be reproduced @Ioannidis2009-hv, which brings into question the reliability of those
studies.
Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
earlier case study @Baggerly2010-nr.

@@

>>>>

Unreliable research is a waste of resources and time, and can mislead readers to pursue doomed
research directions.
Much has been written on how irreproducibility has reached crisis levels in psychology and life
sciences [@Baker2015-bs; @Begley2012-di].
Indeed in bioinformatics, the situation is grim with a 2009 report showing only a small fraction
of studies able to be reproduced @Ioannidis2009-hv, which brings into question the reliability of
those studies.
Irreproducible bioinformatics has the potential to place patient safety at risk, as demonstrated by an
earlier case study @Baggerly2010-nr.

@@

## Literate programming

A common recommendation for reproducibility is to complete all works with computer scripts
and remove the need for any manual steps.
Computer programs (scripts) are preferred over graphical interface tools as the script contains
a complete instruction set for how a piece of work was conducted.
This allows auditing and reproduction which would be difficult for graphical tools like spreadsheets.
Scripts also allow the work to be completely automated and reduce the chance for human error during
any manual steps.
On the other hand, data cleaning and curation is sometimes required, and so manual steps are unavoidable
in some situations.

Literate programming combines "chunks" of analytical code together with human readable text @Knuth1984-di.
After compilation, the resulting output contains the code with computational results
(eg: figures and tables) along with contextualising explanations and narratives.

The Sweave project, concieved in 2001, was designed to give LaTeX documents embedded R code chunks and
was envisaged as a way to generate R documentation materials and generate statistical analysis reports
@Leisch2002-yc.
Around 2015, R Markdown emerged as a more popular alternative, as formatting is simpler as compared to Sweave,
saving time.
Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
R, Python, Julia and other computer language chunks into documents @Perkel2018-rz.
In 2022, the company behind RStudio (Posit) released Quarto, the conceptual successor of R Markdown, but with
enhanced support for other computer languages like Python, Julia, and Observable JavaScript
(and more attractive reports as well) @Bauer2023-eh.

Whether Rmarkdown, Jupyter or Quarto, literate programming offers some substantial benefits over alternative
approaches:

* The provenance of any result can be demonstrated.
The resulting document produced contains the code executed, together with the results (eg: a chart).
This is in contrast to a data analysis report assembled in a word processor, which due to the many
copy-paste operations required might contain errors.

* It saves time.
For the case where a report needs to be run routinely, this would involve significant copy-paste to
be assembled "manually" with a word processor.
Using a literate script would mean only minimal changes are required between interations.

* Literate programs accommodate extensive documentation.
This allows the analyst to include rich descriptions of scientific works.
For example, a report may contain background, methods, results, discussion and references.
Not only text, but various objects can be included like links to other resources,
tables, images, videos, etc.
More information on documentation below.

* Outputs are arranged.
When using a regular script, an analyst might make dozens of charts which are written to the
disk with names like "chart1.png" or "model8.svg".
When the number of outputs is large, it becomes unwieldy and hard to place which plot corresponds
to which part of the script.
By arranging the outputs (eg: charts and tables) in a document in the sequence that they were generated,
it helps the reader understand the logical steps taken in any analysis.

* Literate scripts are amenable to version control.
Version control is a useful best practice in software development (discussed below).

* The resulting document is only rendered when the entire script has compiled without errors.
In contrast, a regular script might generate some outputs and then encounter an error,
so we are not sure whether the whole script is free of errors.

* Flexible output formats.
These include PDF, DOC/DOCX and HTML, with HTML having some notable benefits to other formats
including better usability for mobile devices like phones and tablets, and the ability to
support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
